---
title: "Base R, data.table and tidyverse"
author: "George Savva"
date: "22/02/2022"
output:
  html_document:
    toc: yes
    toc_depth: 2
---

# Limitations of "base" R

With only base R (that is, R without add on packages) it can be unexpectedly difficult to perform some simple tasks.

A good example is making a table of summary statistics.  This is difficult with base R but is simple with using function from add-on packages. 

Here I illustrate this using two widely used systems for data manipulation in R, namely `data.table` and `tidyverse`.  Finally I illustrate a new package, `gtsummary` that is specifically designed for creation of publication ready summary tables.

# Mean of one variable stratified by another

Suppose we have a dataset of the heights (in cm) of 100 men and women, and we want to make a descriptive table of means, standard deviations and counts by sex.

First let's make a fake dataset.  We'll assume women have an average height of 170cm, men of 180cm with both groups normally distributed with a standard deviation of 10.

```{r }
# Look up 'sample' to understand what this does
sex = sample(c("Male", "Female"), size=100, replace=TRUE)

# What does this line do?
height = rnorm(n=100, 
               mean=180 + 10* (sex=="Male"), 
               sd=10)

# We have two vectors of the same length so we can combine them into a data frame.
dat <- data.frame(height,sex)

```

Now lets quickly check the data using a plot to make sure it looks as we would expect.

```{r }
head(dat)
boxplot(dat$height~dat$sex)
```


# The base R way to get summary statistics

The `aggregate` function can be used to calculate a single statistic over groups as follows:

```{r }
aggregate( height ~ sex , FUN=mean, data=dat)
```

Alternatively we could use `tapply`:

```{r }
tapply( dat$height, dat$sex , FUN=mean )
```

While this works it is difficult to get a more complicated table.  There is no obvious way to get a table of means, standard deviations and counts without using an external package.

I'll illustrate three different approaches of here.  `data.table` and `tidyverse` add ways to manipulate data

It is likely that if you spend a lot of time using R you should learn to use one or other (or both) of these systems.

Finally I'll illustrate `gtsummary`, a package specifically designed to make tabulation of results easier.


# Using `data.table`

The `data.table` package gives us a very flexible way to perform fast grouped operations on datasets.  A `data.table` is an enhanced version of a `data.frame`, and the main function of `data.table` an extension to the `[ ]` operator (square brackets) that is much more powerful than the default R version.

First we need to load the package, then turn the data.frame into a data.table using `setDT`.

```{r }
library(data.table)
setDT(dat)
```

Then we can use the extended square bracket syntax to create our table.

First the simple comparison of means:

```{r }
dat[ , mean(height), by=sex ]
```

Next, adding the counts and standard deviations:

```{r }
dat[ , .(Count=.N, Mean=mean(height),SD=sd(height)), by=sex]
```

## The data.table syntax

The `[` operator in data.table has three arguments.  In short, we express a command on a dataset (here called `dat`) by specifying:

```{r eval=FALSE}
dat[ which rows to use , what to do , which columns to group on ]
```

In the first version of the command above we left the first entry blank (so used all the rows), placed `mean(height)` in the second position and specified `by=sex` in the third. In the second version we expanded the second argument to return a list of elements, and gave them new names.

For more details of using data.table, see: https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html

# The `tidyverse` way

`tidyverse` is a set of `R` packages that provide many functions for data manipulation and programming.  In particular the `dplyr` library includes functions for data manipulation and summarisation.  To use these we can load the `dplyr` library:

```{r }
library(dplyr)
```

Then get our results table the `tidyverse` way:

First just the means

```{r }
dat %>% group_by(sex) %>% summarise(mean(height))
```

Now with the counts and standard deviations:

```{r  message=FALSE, warning=FALSE}
dat %>% group_by(sex) %>% summarise(N=n(),mean(height), sd(height))
```

## The tidyverse syntax

`dplyr` introduces six main functions for manipulating and summarising data, these are `mutate`, `arrange`, `select`, `filter`, `summarise`, and `group_by`.  Using combinations of these functions you can perform most simple data operations.  Functions are chained together using the pipe operator `%>%` which passes the output from one into the next.  So the first command above reads something like:  "take dat, then group it by sex, then for each group return the summary statistics we specified".   

Visit https://www.tidyverse.org/learn for more.

# Using `gtsummary`

Finally to illustrate a package meant specifically for nicely formatted data tabulations, the `tbl_summary` function from the `gtsummary` package can create tables of descriptives.

```{r message=FALSE, warning=FALSE}
library(gtsummary)
tbl_summary(dat, by=sex, statistic=list(height~"{mean} ({sd})"))
```



# Which to use?

`data.table` and `tidyverse` perform a lot of the same tasks in improving the `R` experience.  You will probably choose to mainly use one or the other (or to mostly stick with base R), but it will help to be familiar with both if you want to understand and reuse code written by others.   

If you search `tidyverse` vs `data.table` online you will find a lot of differing opinions as to which to use.  More people learn `tidyverse` now, possibly because it has a lot of resources put into its development and promotion.  Personally I like the `data.table` syntax better so I tend to use this, borrowing from `tidyverse` packages when I need to.  

I find that the more confident I get with it, the more `data.table` features I use, which I think improves my code.

